Computing the {\em least-commitment} action model from state observations

### 2018-12-24 doc/ijcai-19-conformant/conformant-learning.tex
adding updates to the IJCAI19 paper on conformant planning

\section{Introduction}
Given an input sequence of partially observed states, this paper formalizes the task of computing the {\em least-commitment} action model that is able to {\em explain} the given observation. This task is of interest because it allows the incremental learning of action models from arbitrary large sets of partial observations.

@@

\section{Computing the {\em least-commitment} action model from observations}
First, this section formalizes the notion of the {\em least-commitment} action model that is able to {\em explain} an observation. Next, the section describes our approach to compute such model.

\subsection{The {\em least-commitment} action model}
The task of computing the {\em least-commitment} action model from a sequence of state observations is defined as $\tup{\Phi,\mathcal{O}(\tau)}$:
\begin{itemize}
\item $\Phi=\tup{F,A[\cdot]}$ is a {\em classical planning frame} where the semantics of each action $a\in A[\cdot]$ is unknown; i.e. the corresponding $\tup{\rho,\theta}$ functions are undefined. We say that an {\em action model} $\mathcal{M}$ is a definition of the $\tup{\rho,\theta}$ functions of every action in $A[\cdot]$. 
\item $\mathcal{O}(\tau)$ is a sequence of partial states that results from the partial observation of a trajectory $\tau(\pi,s_0)$ within the {\em classical planning frame} $\Phi$.
\end{itemize}

@@

\subsection{{\em Partially specified} \strips\ action models}
A set of action models can be defined {\em explicitly}, by enumerating all the models that belong to the set or {\em impolitely}, by enumerating the constraints that must satisfy any model that belongs to the set. Inspired by the notion of {\em incomplete (annotated) model}~\cite{sreedharan2018handling}, we introduce here {\em partially specified} \strips\ action models, a formalism for the {\em implicit} representation of a set of \strips\ schema.

We show now that extending to the {\em knowledge level} our propositional encoding of \strips\ action schemes, we can compactly represent a set of \strips\ schema. The extension defines, for each proposition {\tt\small $pre\_e\_\xi$}, with two propositions {\tt\small $K\pre\_e\_\xi$} and {\tt\small $K\neg\pre\_e\_\xi$}, meaning that is {\em known} that $e\in{\mathcal I}_{\Psi,\xi}$ is a precondition of $\xi$ and that is {\em known} that $e\in{\mathcal I}_{\Psi,\xi}$ is not a precondition of $\xi$. Likewise for each {\tt\small $eff\_e\_\xi$} proposition we define these two, {\tt\small $K\eff\_e\_\xi$} and {\tt\small $K\neg\eff\_e\_\xi$}, meaning that is {\em known} that $e\in{\mathcal I}_{\Psi,\xi}$ is an effect of $\xi$ and that is {\em known} that $e\in{\mathcal I}_{\Psi,\xi}$ is not an effect of $\xi$. With respect to this formalization, the {\em full space} of possible \strips\ schemas for $\xi$ is compactly represented as {\tt\small $\bigcap_{e\in{\mathcal I}_{\Psi,\xi}} \neg K\pre\_e\_\xi\wedge \neg K\neg\pre\_e\_\xi \wedge\neg K\eff\_e\_\xi\wedge \neg K\neg\eff\_e\_\xi$}.

Figure~\ref{fig:partial} shows a {\em partially specified} \strips\ model that represents of a set of action models for the {\tt\small stack(?v1,?v2)} schema. This {\em partially specified} \strips\ model compactly represents a set of $2^5$ models since (1), the actual five effects of the action are {\em unknown}, (2) preconditions are known to appear as in the actual definition of the {\tt\small stack(?v1,?v2)} schema for the {\em blocksworld} and (3), it is {\em known} that the preconditions and effects that are not part of the actual definition of the {\tt\small stack(?v1,?v2)} schema are not part of this {\em partially specified} \strips\ model.

@@

\subsection{Computing the {\em least-commitment} model via conformant planning}
The compilation for learning \strips\ action models defined by~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning} can be used to compute a {\em partially specified model} that explains a given observation by producing a conformant planning task instead of a classical planning task. To do so the initial state $\Upsilon$ of the conformant planning task does not code the {\em empty} action model but the full space of action models. In more detail, the clauses in $\Upsilon$ comprises:
      \begin{enumerate}
      \item The {\em unit clauses} given by the fluents that hold in the initial state $I=s_0$.
      \item The clauses representing that the actual value of fluents {\tt\small pre\_e\_$\xi$}, {\tt\small eff\_e\_$\xi$} is unknown. In other words, that any model from the \strips\ space of models (following the previously mentioned syntactic constraints) can initially be part of the {\em least-commitment} action model. Formally, for every $\xi$ and $e\in{\mathcal I}_{\Psi,\xi}$, then $\Upsilon$ includes these two clauses:
            \begin{itemize}
            \item {\tt\small pre\_e\_$\xi$} $\vee$ {\tt\small $\neg$pre\_e\_$\xi$}.
            \item {\tt\small eff\_e\_$\xi$} $\vee$ {\tt\small $\neg$eff\_e\_$\xi$}.
            \end{itemize}
            One can also add here clauses that encode {\em domain-specific constraints} (as mentioned in the previous section) to make the conformant planning problem easier for a specific domain.
      \end{enumerate}
A conformant plan that solves this task is a {\em partially specified model} that {\em explains} the input observation. To compute the {\em least-commitment} action model, we have to associate a cost of 1 to the actions that {\em program} a given precondition or effect of an action schema, a cost of 0 to the remaining actions, and compute the {\em optimal} plan for the resulting conformant planning task.

@@

>>>>

\section{Introduction}
Given an input sequence of partially observed states, this paper formalizes the task of computing the {\em least-commitment} action model that is able to {\em explain} that input observation.

This task is of interest because, if we can keep the set of all the action models that are consistent with the observation, i.e. the {\em version space}~\cite{mitchell1982generalization}, then each new observation has either no effect on the {\em version space} or eliminates some models from it. This allows to implement an incremental prunning of the {\em version space} and hence, to learn planning action models from arbitrary large sets of partial observations (one has never to go back and re-process old observations).

@@

\section{The {\em least-commitment} action model}
The task of computing the {\em least-commitment} action model from a sequence of state observations is defined as $\tup{\Phi,\mathcal{O}(\tau)}$:
\begin{itemize}
\item $\Phi=\tup{F,A[\cdot]}$ is a {\em classical planning frame} where the semantics of each action $a\in A[\cdot]$ is unknown; i.e. the corresponding $\tup{\rho,\theta}$ functions are undefined. We say that an {\em action model} $\mathcal{M}$ is a definition of the $\tup{\rho,\theta}$ functions of every action in $A[\cdot]$. 
\item $\mathcal{O}(\tau)$ is a sequence of partial states that results from the partial observation of a trajectory $\tau(\pi,s_0)$ within the {\em classical planning frame} $\Phi$.
\end{itemize}

@@

\subsection{{\em Partially specified} \strips\ action models}
A set of action models can be defined {\em explicitly}, enumerating all the models that belong to the set or {\em implicitly}, enumerating the constraints that must satisfy any model that belongs to the set. Inspired by the notion of {\em incomplete (annotated) model}~\cite{sreedharan2018handling}, we introduce here {\em partially specified} \strips\ action models, a formalism for the {\em implicit} representation of a set of \strips\ schema.

We show now that extending to the {\em knowledge level} our propositional encoding of \strips\ action schemes, we can compactly represent a set of \strips\ schema. The extension defines, for each proposition {\tt\small $pre\_e\_\xi$}, two propositions {\tt\small $K\pre\_e\_\xi$} and {\tt\small $K\neg\pre\_e\_\xi$}, meaning that is {\em known} that $e\in{\mathcal I}_{\Psi,\xi}$ is a precondition of $\xi$ and that is {\em known} that $e\in{\mathcal I}_{\Psi,\xi}$ is not a precondition of $\xi$. Likewise for each {\tt\small $eff\_e\_\xi$} proposition we define these two, {\tt\small $K\eff\_e\_\xi$} and {\tt\small $K\neg\eff\_e\_\xi$}, meaning that is {\em known} that $e\in{\mathcal I}_{\Psi,\xi}$ is an effect of $\xi$ and that is {\em known} that $e\in{\mathcal I}_{\Psi,\xi}$ is not an effect of $\xi$.

\begin{definition}[Partially specified model]
A {\em partially specified action schema} $\xi[\cdot]$ is an aritrary formula over the {\tt\small $K\pre\_e\_\xi$}, {\tt\small $K\neg\pre\_e\_\xi$}, {\tt\small $K\eff\_e\_\xi$} and {\tt\small $K\neg\eff\_e\_\xi$} propositions ($e\in{\mathcal I}_{\Psi,\xi}$) whose valuations express the enumeration of the set of action models $M_\xi$.
\end{definition}

With respect to this formalization, the {\em full space} of possible \strips\ schemas for $\xi$ is compactly represented by the {\em partially specified action schema} {\tt\small $\bigcap_{e\in{\mathcal I}_{\Psi,\xi}} \neg K\pre\_e\_\xi\wedge \neg K\neg\pre\_e\_\xi \wedge\neg K\eff\_e\_\xi\wedge \neg K\neg\eff\_e\_\xi$}. Figure~\ref{fig:partial} shows a {\em partially specified} \strips\ model that represents of a set of action models for the {\tt\small stack(?v1,?v2)} schema. This {\em partially specified} \strips\ model compactly represents a set of $2^5$ models since (1), the actual five effects of the action are {\em unknown}, (2) preconditions are known to appear as in the actual definition of the {\tt\small stack(?v1,?v2)} schema for the {\em blocksworld} and (3), it is {\em known} that the preconditions and effects that are not part of the actual definition of the {\tt\small stack(?v1,?v2)} schema are not part of this {\em partially specified} \strips\ model.

@@

\section{Incremental learning of action models from observations}
This section sows how to implement an incremental learning of action models by prunning action models that are not consistent with the input observations.

\subsection{Computing the {\em least-commitment} model via conformant planning}
The compilation for learning \strips\ action models defined by~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning} can be used to compute a {\em partially specified model} that explains a given observation by producing a conformant planning task instead of a classical planning task. The initial state $\Upsilon$ of the conformant planning task does not code an {\em initial} action model (typically {\em empty}) but the full space of action models. In more detail, the clauses in $\Upsilon$ comprises:
      \begin{enumerate}
      \item The {\em unit clauses} given by the fluents that hold in the initial state $I=s_0$.
      \item The clauses representing that the actual value of fluents {\tt\small pre\_e\_$\xi$}, {\tt\small eff\_e\_$\xi$} is unknown. In other words, that any model from the \strips\ space of models (following the previously mentioned {\em syntactic constraints}) can initially be part of the {\em least-commitment} action model. Formally, for every $\xi$ and $e\in{\mathcal I}_{\Psi,\xi}$, then $\Upsilon$ includes these two clauses:
            \begin{itemize}
            \item {\tt\small pre\_e\_$\xi$} $\vee$ {\tt\small $\neg$pre\_e\_$\xi$}.
            \item {\tt\small eff\_e\_$\xi$} $\vee$ {\tt\small $\neg$eff\_e\_$\xi$}.
            \end{itemize}
            One can also add here clauses that encode {\em domain-specific constraints} (as mentioned in the previous section) to make the conformant planning problem easier to be solved for a specific domain.
      \end{enumerate}
A conformant plan that solves this task induces a {\em partially specified model} that {\em explains} the input observation. To get the {\em least-commitment} action model, we have to compute the {\em optimal} plan for the resulting conformant planning task when: The actions that {\em program} a given precondition (or effect) of an action schema have an associated cost of 1. These actions represent now a reduction of one unit in the {\em uncertainty} of the initial {\em partially specified model}. In other words, these actions implement the different possible {\em immediate specification} of the current {\em partially specified model}. The remaining actions have a cost of 0.

@@

\subsection{Learning algorithm}
Given a {\em classical planning frame} $\Phi=\tup{F,A[\cdot]}$ where the semantics of each action $a\in A[\cdot]$ is unknown and a set of state observations $\mathcal{O}_1,\ldots,\mathcal{O}_T$. Our learning algorithm for the incremental learning of action models from state observations is defined as follows:
\begin{enumerate}
\item Build the {\em least-commitment} task $\tup{\Phi,\mathcal{O}_t}$ to process observation $\mathcal{O}_t$, {\tt\small $1\leq t\leq T$}.
\item Build the classical planning task $P=\tup{F',A',I',G'}$ to solve the previous task.
\item Extract the {\em partially specified} model from the previous solution.
\item Continue until, the {\em version space} is a singleton or no more observations left
\end{enumerate}

Assuming and a set of independent observations $\{\mathcal{O}_1,ldots,\mathcal{O}_t\}$ we need to see at least $\frac{ln{1-\delta} + ln{}}{\epsilon}$ to be sure that the prediction error is at most $\epsilon$ with probability $(1-\delta)$.
